---
title: "Practical Machine Learning - Course Project"
author: "Daniel Harcek"
date: "20 March 2015"
output:
  md_document:
    variant: markdown_github
  html_document:
    theme: united
---

## Synopsis
#### predmachlearn-012 // 6th of April 2015

Predict activity quality class (see below) based on data gathered from accelerometers on the belt, forearm, arm, and dumbell.

You should

* create a report describing how you built your model, 
* how you used cross validation, 
* what you think the expected out of sample error is, 
* and why you made the choices you did. 

You will also use your prediction model to predict 20 different test cases. 

## Description

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. **One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it**. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts **correctly and incorrectly in 5 different ways**. More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset). 

The training data for this project are available here: 
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>

The test data are available here:
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>

Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions:

* exactly according to the specification (Class A)
* throwing the elbows to the front (Class B)
* lifting the dumbbell only halfway (Class C)
* lowering the dumbbell only halfway (Class D)
* throwing the hips to the front (Class E).

Read more: <http://groupware.les.inf.puc-rio.br/har#ixzz3UxT89OCG>

## R-Part

### Load necessary libraries

We load both `rpart` and `randomForest` because as we will later see, regular decision tree performs with low prediction precission so we experiment instead with `randomForest`.

```{r load_libraries}
library("caret")
library("corrplot")
library("rpart")
library("rattle")
library("rpart.plot")
library("randomForest")
```

### Download and load the data

We download the data if they are not present in working directory and load datasets into the memory.

```{r download_and_read_data}
URL_BASE <- "https://d396qusza40orc.cloudfront.net/predmachlearn/"
DATA_TRAINING <- "pml-training.csv"
DATA_TESTING <- "pml-testing.csv"
if (!(file.exists(DATA_TRAINING))) {
  download.file(url=paste(URL_BASE, DATA_TRAINING, sep=""), destfile=DATA_TRAINING, method="curl")
  download.file(url=paste(URL_BASE, DATA_TESTING, sep=""), destfile=DATA_TESTING, method="curl")
}

activity_training <- read.csv(file=DATA_TRAINING, head=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""))
activity_test <- read.csv(file=DATA_TESTING, head=TRUE, sep=",", na.strings=c("NA","#DIV/0!",""))
```

### Preparing the dataset

We start exploring the data. We see that the number of the predictors is quite hi, but looking at the summary and complete cases we can see that there is a lot of columns with high number of NA's.

```{r basic_exploration}
ncol(activity_training)
str(activity_training)
summary(activity_training) # remove data with a lot of NA's
table(complete.cases(activity_training))
```

We will all such data (majority of NA's in predictor observations) and we remove also variables giving no information in relation to predicted class as a timestamp, user name etc.

```{r making_data_tidy}
activity_training_wo_na <- activity_training[,colSums(is.na(activity_training)) < 19000]
table(complete.cases(activity_training_wo_na))
nrow(activity_training_wo_na)
ncol(activity_training_wo_na)

activity_training_tidy <- activity_training_wo_na[ ,!(names(activity_training_wo_na) %in% c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window"))]
ncol(activity_training_tidy)
```

We see that we successfully reduced potential number of predictors from `162` to `52` (if we omit the class value itself). We will further explore predictors having zero variance and highly correlated predictors.

```{r explore_potential_zero_variance}
nearZeroVar(subset(activity_training_tidy, select=-c(classe)), saveMetrics=TRUE)
```

It seems that there are no zero variance predictors, let's move on searching for highly correlated data. We compute the correlation matrix and quickly visually inspect the correlation plot. 

```{r highly_correlated_variables}
cormatrix <- cor(subset(activity_training_tidy, select=-c(classe)))
corrplot(cormatrix, order = "hclust", type = "lower", main="Dumbbell Biceps Curl Activity Tracker Data", tl.pos="n")
```

We can see that there is couple of predictors with correlation over level 9 (which is a common default value for cutoff), so we will remove these.

```{r cut_correlated}
cutcorrelated <- findCorrelation(cormatrix, cutoff=.90, verbose=FALSE)
length(cutcorrelated)
activity_training_predictors <- activity_training_tidy[,-cutcorrelated]
dim(activity_training_predictors)
```

### Training and testing

We first split the data to training and test set.

```{r create_training_and_data_set}
set.seed(8)
trainIndex <- createDataPartition(activity_training_predictors$classe, p=.60, list=FALSE)
train <- activity_training_predictors[trainIndex, ]
test <- activity_training_predictors[-trainIndex, ]
```

Finally we can train a simple classification tree on our data.

```{r train_tree}
fit <- train(classe ~ ., data=train, method="rpart")
```

And review properties of our tree. 

```{r check_trained_tree}
print(fit$finalModel)
fancyRpartPlot(fit$finalModel)

prediction <- predict(fit, newdata=test)
confusionMatrix(prediction, test$classe)$overall
```

And we see that accuracy of trained decision tree is low. 
We try to improve this preprocessing the data with centering, scaling and doing 3 fold cross validation.

```{r train_tree_preprocess_and_3fold_cv}
fit <- train(classe ~ ., data=train, method="rpart",
              preProcess=c("center", "scale"),
              trControl=trainControl(method="cv", number=3, allowParallel=TRUE))

fit
prediction <- predict(fit, newdata=test)
confusionMatrix(prediction, test$classe)
confusionMatrix(prediction, test$classe)$overall
```

This did not bring any improvement at all. 
Let's try random forest. Couple of observations here, I ended up using `randomForest` function since the `train` function from `caret` package was extremely slow. Some posts suggested to use `parRF` method in `caret` to speed things up, but it did not help. I did not used formula form as [advised here](http://stats.stackexchange.com/a/37382) and based on OOB (out of bag) error tracing decided seeing that OOB almost bottomed down around` `ntree` equal 300 to lower down the `ntree` value from `500` (OOB 0.87%) to `300` (OOB 0.91%).

```{r train_random_forest}
predictors <- subset(train, select=-c(classe))
decision <- subset(train, select=c(classe))
fit <- randomForest(predictors,decision[,1], prox=TRUE, ntree=300) # drop empty dimension https://stat.ethz.ch/pipermail/r-help/2008-July/166617.html

prediction <- predict(fit, newdata=test)
confusionMatrix(prediction, test$classe)$overall
```

### Final Out Of Sample Error

We see that `randomForest` improved out-of sample accuracy dramatically from `0.49` to `0.99` with trading off the computation time for precision. That means our out of sample error is in the end `0.001`.

### Prediction Assignment Submission

Let's now remove all unused predictors from test data, generate predictions and prepare the submissions.
```{r predict_test_data_and_prepare_to_submit_answers}
activity_test_cleaned <- activity_test[names(activity_test) %in% colnames(activity_training_predictors)]
prediction <- predict(fit, newdata=activity_test_cleaned)
prediction

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(prediction)
```
